\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-10-07 aistats/sample_paper.tex
preparing submission for aistats

\documentclass[twocolumn,superscriptaddress,aps]{revtex4-1}

\usepackage[utf8]{inputenc}

@@

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

@@

\title{\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}}
\vspace{1cm}
\author{\small{\bf Gilles Louppe}\thanks{\texttt{g.louppe@nyu.edu}}}
\affiliation{New York University}
\author{\small{\bf Kyle Cranmer}\thanks{\texttt{kyle.cranmer@nyu.edu}}}
\affiliation{New York University}

\begin{abstract}

Complex computer simulators are increasingly used across fields of science as
generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely admit a tractable density or likelihood
function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
We adapt the training procedure of generative
adversarial networks by replacing the differentiable generative network with a
domain-specific simulator. We solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the two adversarial objectives.
Effectively, the procedure results in learning a
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
We present results of the method with simulators producing both discrete and continuous data.

\end{abstract}

\maketitle

@@

In parallel, with the introduction of variational
auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While some of these models
also do not admit a tractable density, they are all differentiable by construction.
In addition, generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

@@

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}\label{eq:marginal_likelihood}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown mean.
             ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
             ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
             ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

@@

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The architecture of the critic is the same as in the previous example.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
    \caption{Electron--positron annihilation.
    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
                     Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
                     The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:weinberg}
\end{figure}

@@

A typical physics simulator for this process includes a more precise treatment of the
quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
using \texttt{MadGraph}~\citep{Alwall:2011uj},  ionization of matter in the
detector due to the passage of the out-going $\mu^+\mu^-$ particles using
\texttt{GEANT4}~\citep{Agostinelli:2002hh}, electronic noise and other details of the sensors
that measure the ionization signal, and the deterministic algorithms that
estimate the polar angle $A$ based on the sensor readouts. The simulation of
this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.

@@

This work sits at the intersection of several lines of research related to
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative networks, and variational inference.
Viewed from the literature around implicit generative models based on neural networks,
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and to
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.
In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
are alternatives to adversarial learning that are worth considering.

More generally, likelihood-free
inference is intimately tied to a class of algorithms that can be framed as
density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
inference algorithms are formulated as an iterative two-step process where the
model distribution is first compared to the true data distribution and then
updated to make it more comparable to the latter. Relevant work in this
direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
\citep{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\citep{meeds2015hamiltonian}
estimate gradients with respect
to $\bftheta$ through finite differences from multiple forward passes of the
simulator with variance reduction strategies based on controlling the source of
randomness used for the latent variable $\bfz$.

>>>>

\documentclass[twoside]{article}
\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Adversarial Variational Optimization of\\
              Non-Differentiable Simulators}

\aistatsauthor{Author 1 \And Author 2}
\aistatsaddress{Institution 1 \And Institution 2}

% \aistatsauthor{Gilles Louppe \And Kyle Cranmer}
%
% \aistatsaddress{New York University\\
%                 \& University of Li{\`e}ge
%                 \And New York University}
                ]

\begin{abstract}
    Complex computer simulators are increasingly used across fields of science as
    generative models tying parameters of an underlying theory to
    experimental observations. Inference in this setup is often
    difficult, as simulators rarely admit a tractable density or likelihood
    function. We introduce Adversarial Variational Optimization (AVO), a likelihood-free
    inference algorithm for fitting a non-differentiable generative model incorporating ideas from empirical Bayes and variational inference.
    We adapt the training procedure of generative
    adversarial networks by replacing the differentiable generative network with a
    domain-specific simulator. We solve the resulting non-differentiable
    minimax problem by minimizing variational upper bounds of the two adversarial objectives.
    Effectively, the procedure results in learning a
    proposal distribution over simulator parameters, such that the corresponding
    marginal distribution of the generated data matches the observations.
    We present results of the method with simulators producing both discrete and continuous data.
\end{abstract}

@@

In parallel, with the introduction of variational
auto-encoders~\cite{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\cite{2016arXiv161003483M}.  While some of these models
also do not admit a tractable density, they are all differentiable by construction.
In addition, generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

In this note, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of both algorithms is to match marginal distribution of
the generated data to the empirical distribution of the observations.

@@

As thoroughly explored in \cite{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\cite{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\equiv \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \cite{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\cite{ranganath2014black, tran2017variational}.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}\label{eq:marginal_likelihood}
\qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \cite{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
\mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

We can view the optimization of the marginal likelihood $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

@@

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
\caption{Discrete Poisson model with unknown mean.
 ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
            the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density.
 ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:poisson}
\end{figure}

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\cite{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
semi-positive definite $5 \times 5$ matrix defining a fixed projection
of $\bfz$ into the observed space.
\end{itemize}
Again, the AVO algorithm does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The architecture of the critic is the same as in the previous example.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
\caption{Electron--positron annihilation.
({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
         Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
         The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:weinberg}
\end{figure}

@@

A typical physics simulator for this process includes a more precise treatment of the
quantum mechanical  $e^+e^- \rightarrow \mu^+\mu^-$ scattering
using \texttt{MadGraph}~\cite{Alwall:2011uj},  ionization of matter in the
detector due to the passage of the out-going $\mu^+\mu^-$ particles using
\texttt{GEANT4}~\cite{Agostinelli:2002hh}, electronic noise and other details of the sensors
that measure the ionization signal, and the deterministic algorithms that
estimate the polar angle $A$ based on the sensor readouts. The simulation of
this process is highly non-trivial as is the space of latent variables $\mathcal{Z}$.

@@

This work sits at the intersection of several lines of research related to
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative networks, and variational inference.
Viewed from the literature around implicit generative models based on neural networks,
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A, 2017arXiv170107875A} to
non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization}.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\cite{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and to
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.
In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions~\cite{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
are alternatives to adversarial learning that are worth considering.

More generally, likelihood-free
inference is intimately tied to a class of algorithms that can be framed as
density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
inference algorithms are formulated as an iterative two-step process where the
model distribution is first compared to the true data distribution and then
updated to make it more comparable to the latter. Relevant work in this
direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
\cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
estimate gradients with respect
to $\bftheta$ through finite differences from multiple forward passes of the
simulator with variance reduction strategies based on controlling the source of
randomness used for the latent variable $\bfz$.

Likewise, AVO closely relates to recent extensions of GANs, such as
Adversarial Learned Inference (ALI)~\cite{dumoulin2016adversarially},
Bidirectional GANs (BiGANs)~\cite{donahue2016adversarial},
$\alpha$-GAN~\cite{rosca2017variational}, Adversarial Variational Bayes
(AVB)~\cite{DBLP:journals/corr/MeschederNG17}, and the \texttt{PC-Adv}
algorithm of  \cite{2017arXiv170208235H}, which
add an inference network to the generative model.  Each of these assume a
tractable density $p(\bfx|\bftheta)$ that is differentiable with respect to
$\bftheta$, which is  not satisfied in the likelihood-free setting. Our lifting
of the non-differentiable simulator $p(\bfx|\bftheta)$ to the variational
program $q(\bfx | \bfpsi)$ provides the ability to differentiate expectations
with respect to $\bfpsi$ as in Eqn~\ref{eqn:approx-grad}; however, the density
$q(\bfx | \bfpsi)$ is still intractable. Moreover, we do not attempt to define a
recognition model  $q(\bfz, \bftheta|\bfpsi)$ as the latent space $\mathcal{Z}$
of many real-world simulators is complicated and not amenable to neural a
recognition model.
